Allee Effect and Cultural Evolution

### 2015-05-11 manuscript/draft.tex
Updated Draft
-This is a temp. update so I can work from home in the weekend
Updated manuscript

We investigate the equilibrium conditions for different initial population sizes of  $a$ and $b$ and different settings of $z$, $A$, and $K$, assuming that the $r_a=r_b$, and for convenience holding the condition $K_b>K_a$. Given that equations \ref{eq1},~\ref{eq4} have no analytical solution, we identified equilibrium conditions via simulation (see appendix 1 for details; source codes can be found on the following repository %put link here
as well as in the ESM).  
%if the points we need to reiterate are short perhaps a footnote will suffice instead of an appendix?

@@

\subsection{Effect of social learning $z>0$}

\subsection{Effect of competition}

\section{Experiments}

We investigate different initial conditions of $m$ and $n$ for 6 different values of z (0,0.05,0.1,0.2,0.3, and 0.5). For each parameter combination we apply equations (\ref{eq1},~\ref{eq4}) for 50,000 iterations and record the final values of $m$ and $n$, identifying the following equilibria.

\enrico{We might need to use a rounding factor. And in fact these introduces to the problem of scale and difference between discrete and continuos  models.}
\enrico{We can analytical solve equations 1~4 to shows that the E1, E2 and E3 are indeed equilibria. Perhaps in an appendix? (although this is quite obvious...)}

We are aware that condition U can potentially become one of the other stable equilibria after the 50,000 simulations, hence we computed the Lyapunov exponent to define the stability of the system. We devised three core scenarios reflecting fixing b=2 and equating Rn to gRm, where g is given by the following equation:

\xavi{equation5, lyapunov, where this does go?}

so that when Rn=gRm,  Rmax(m,t) and Rmax(n,t) becomes equal. In all scenarios we considered Allee thresholds and carrying capacities as proportions by ensuring that max(Km,Kn)=1.

\enrico{Justify choice of arbitrary values}

\subsection{Scenario 1: Traits with identical payoffs (i.e. Am = An   and Km=Kn)}

The first scenario assumes two traits that are mathematically equivalent, so that $A_m = A_m, K_m=K_n and r_m=r_n$. Figure 1 shows the basin of attractions of each of the four equilibria for different values of z, setting r=0.005, .A=0.2 and K=1. When the z=0, i.e. when cultural transmission is disabled we have straight boundaries between the basins of E1, E2m, E2n, and E3 defined by Allee threshold. In other words when both one or both the population are below A we have some extinction events (E3 when mt=0<Am and nt=0<An, E2m when mt=0>Am and nt=0<An and E3n when mt=0<Am and nt=0>An). When z>0, the boundaries of these basins changes and with high z we see the emergence of unstable equilibria (U).

In all cases the E1 equilibria is maintained when $mt=0=nt=0 > Am=An$ as in this scenario differences arises only when m DIFFERENT OF n, and when the two populations are equally sized, C(m,n) is equal to 0 (see equation [3]).

\subsection{Scenario 2: Overlapping payoffs (i.e.  Am <  An  < Km < Kn)}

Am=0.2, An=0.5, Km=0.7,Kn=1

\subsection{Scenario 3: Internal overlap  (i.e.  Am <  An < Kn <  Km )}

An=0.2, Am=0.5, Km=0.7,Kn=1

\subsection{A note on competition}

@@

\xavi{equation6}

@@

\section{Conclusion}

BlaBlaBla

>>>>

We investigate the equilibrium conditions for different initial population sizes of  $a$ and $b$ and different settings of $z$, $A$, and $K$, assuming that  $r_a=r_b$, and for convenience holding the condition $K_b \geq K_a$. Given that equations \ref{eq1},~\ref{eq4} have no analytical solution, we identified equilibrium conditions via simulation (see appendix 1 for details; source codes can be found on the following repository %put link here
as well as in the ESM).  
%if the points we need to reiterate are short, perhaps a footnote will suffice instead of an appendix? 
% 100,000 iterations 
% basin plots resolution

@@

When $z=0$, equations \ref{eq1},~\ref{eq4} \emph{de facto} portray a standard cubic growth model with Allee effect, where the success (or failure) of a given focal trait is simply given by its own density, and the density of alternative cultural traits has no influence. For a generic cultural trait $x$, the model predicts three point attractors at $x=0$ (extinction), $x=A_x$, and $x=K_x$. The equilibrium $x=A_x$ is unstable in all cases, whereas the stability of the node at $x=K_x$ is a function of the per capita growth rate $r_x$. Increasing the latter decreases the stability of fixed points \citep{scheuring_1999}, leading to the transition from point attractors (with equilibrium at $x=K_x$), to limit cycles and chaotic dynamics with islands of stability (see fig. \ref{fig:bifurcationDiagram}). % Not sure if there are strange attractors.  Reference  any  bifurcation analysis on the equation we used. Scheuring 1999 on JTB?

\begin{figure}[h!]
  \centering
      \includegraphics[width=0.7\textwidth]{./figures/figure1.jpg}
  \caption{Bifurcation diagram of a single population model with $A_x=0.2$ and $K_x=0.8$. For $0<r_x<0.6649747$, any initial value of $x$ greater than $A_x$ converge to $K_x$.}
    \label{fig:bifurcationDiagram}
\end{figure}

Here we consider only scenarios where the node $x=K_x$ is stable. In this case the initial value of $x(t=0)$ will determine the ultimate equilibria of the system: extinction, if $x(t=0)<A_x$; unstable equilibria, if $x(t=0)=A_x$; and stable equilibrium, if $x(t=0)>A_x$. Thus in a 2-trait model, the coexistence of both variants is guaranteed only when the initial density of both traits are above their respective Allee threshold, otherwise we should expect that either (i.e. $a(t=0) \geq A_a$ and $b(t=0)<A_b$ or $a(t=0)<A_a$ and $b(t=0) \geq A_b$) or both (i.e. when $a(t=0)<A_a$ and $b(t=0)<A_b$) variants will be ultimately extinct (see fig. \ref{fig:NoTransmissionBasin}).

\begin{figure}[h!]
  \centering
      \includegraphics[width=0.7\textwidth]{./figures/figure2.pdf}
  \caption{Basins of attraction for the four equilibria (mutual extinction: grey; extinction of $a$: blue; extinction of $b$: red; stable coexistence: green). olid lines represent trajectories of the system for different initial conditions, represented by dots. Model parameters: $A_a=0.2$; $K_a=0.7$; $A_b=0.5$; $K_b=1.0$; $r_a=r_b=0.05$}.
    \label{fig:NoTransmissionBasin}
\end{figure}

\subsection{Effects of social learning $z>0$}

When social learning is enabled (i.e. when $z>0$), the stability of the nodes are perturbed by individuals switching from one strategy to another. The minimum implication of these perturbations is a change in the basins of attractions. Initial conditions closer to the boundaries between one basin and another will most profoundly be affected, potentially leading to a complete different equilibria with slightest change in the degree of reliance on social learning. Figure \ref{fig:basicEffectSocialLearning} shows an example of this. When social learning is absent, the initial conditions $i$, $j$, and $k$ respectively result into ...

%Figure showing z=0 and z=something to illustrate the effect of social learning

When the magnitude of social learning is sufficiently high, the system can exhibit exhibit limit-cycles

%Stability nodes remain unchanged for low z, high z determines unstable equilibria
%illustrate examples with different settings
%overlap analysis and changes in basin size
%For stability analysis show that the settings of r guarantees that the system is at a point equilibrium. First Bifurcation Point plot.

\subsection{Effect of competition}
%what if competition
%show new stable points
%show dynamics?

@@

@@
